Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-05-06 paper/tidy-missing-data-paper.Rmd
section two complete - add citation to R language

Features of tidy data were formally discussed in @Wickham2014, and were discussed in terms of their importance for data science by @Donoho2017, and tools for data analysis. Tidy data is defined by @r4ds as:

@@

Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software. This helps the analyst works swiftly and clearly, closing up opportunities for errors. Tidy data principles are general, but are comprehensively implemented in the R programming language, so this paper focusses on R.

@@

Data manipulation is made input- and output-tidy with R packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr]. These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating. **Data reshaping** goes from long to wide formats; **sorting** arranges rows in a specific order; **filtering** removes rows based on a condition; **transforming**, changes existing variables or adds new ones; **aggregating** creates a single value from many values, say, for example, by computing the minimum, maximum, and mean.

@@

Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model. For example in R, y regressed on x and z is: \code{lm(y ~ x + z)}. Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis. For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model. Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs for data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy [@recipes].

@@

Current tools for missing data are messy. Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations. However, these tools suffer the same problems as modelling for imputation: They use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis. The complex, often multivariate nature of imputation methods also makes makes them difficult to represent. Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.

@@

There are many ways each value can be missing, we adopt the notation used in @VanBuuren2012. The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the _missing data model_ $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$. Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter $\psi$, the overall probability of missingness. Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved. In **MAR**, missingness only depends on data observed, not data missing, that is, $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(R = 1 | Y_{obs}, \psi)$. Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness. In **MNAR**, missingness is related to values observed, and unobserved: $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi)$. This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself.  This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena. Visualizations can help assess whether data is MCAR, MAR or MNAR. Imputation is recommended in most cases of missingness dependence.

<!-- (NOTE - are the above last two sentences OK?) -->

>>>>

Features of tidy data were formally described in @Wickham2014, and were discussed in terms of their importance for data science by @Donoho2017, and tools for data analysis. Tidy data is defined by @r4ds as:

@@

Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software. This helps the analyst works swiftly and clearly, closing up opportunities for errors. Tidy data principles are general, but are comprehensively implemented in the R programming language, so this paper focusses on R [@rcore].

@@

Data manipulation is made input- and output-tidy with R packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr]. These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating. Data reshaping goes from long to wide formats; sorting arranges rows in a specific order; filtering removes rows based on a condition; transforming, changes existing variables or adds new ones; aggregating creates a single value from many values, say, for example, in computing the minimum, maximum, and mean.

@@

Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model. For example in R, y regressed on x and z is: \code{lm(y ~ x + z)}. Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis. For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model. Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs into a tidy data format for use data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy [@recipes].

@@

Current tools for missing data are messy. Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations. However, these tools suffer the same problems as modelling: They use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis. The complex, often multivariate nature of imputation methods also makes makes them difficult to represent. Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.

@@

There are many ways each value can be missing, we adopt the notation used in @VanBuuren2012. The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the _missing data model_ $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)$. Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter $\psi$, the overall probability of missingness. Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved. In **MAR**, missingness only depends on data observed, not data missing, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)$. Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness. In **MNAR**, missingness is related to values observed, and unobserved: $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)$. This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself.  This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena. Visualizations can help assess whether data may be MCAR, MAR or MNAR. Imputation is recommended in most cases of missingness dependence.

<!-- (NOTE: Are the above last two sentences OK?) -->
<!-- (NOTE: trying to come up with a sentence that leads from this section to the next) -->
